---
title: "movement in rtsetse"
author: "Andy South"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
#to produce a pdf
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{movement in rtsetse}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

[rtsetse](https://github.com/AndySouth/rtsetse) is an R package that simulates tsetse fly populations on a spatial grid to enable investigation of tsetse control.   

The movement of tsetse flies is an important part of the simulation.   

This vignette demonstrates movement functions of different levels of detail.

First we should load the package after installing from github if necessary.
```{r, eval=TRUE, echo=TRUE, message=FALSE}
#require(devtools)    
#install_github('AndySouth/rtsetse')
require(rtsetse)
```

rtMoveReflectNoGoVeg() simulates movement with reflecting boundaries, avoidance of no-go areas and dependence on vegetation.

This, and the other movement functions in rtsetse, accept a spatial matrix (y,x) defining the population of tsetse flies (of a particular age and gender). The function can be called repeatedly for different ages and sexes, allowing the movement rates to be age and sex dependent. For simplicity the examples shown here will just be for a single age and sex.

### Movement with no vegetation effects or no-go areas

```{r, "movement-no-nogo-or-veg", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#set dimensions of the spatial grid
nY <- 5
nX <- 4
nDays <- 12
#create array to store grids over time
a <- array(0, dim=c(nY, nX, nDays))

#populate central cell of starting grid on day1
a[3,3,1] <- 1
#set proportion moving
pMove <- 0.6

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove)  
}

#quick way of displaying population spread over time
require(raster)
require(sp)
require(RColorBrewer)
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )
```


### Movement avoiding no-go areas

To avoid 'no-go' areas the code is identical to the above with the addition of a matrix for no-go areas. In this example the top row is set as a no-go area. You should be able to see that row remains unoccupied in the spread maps below. 

```{r, "movement-nogo", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#create array to store grids over time
a <- array(0, dim=c(nY, nX, nDays))

#creating empty no-go matrix
mnog <- array(1, dim=c(nY,nX))
#set first row to nogo (0)
mnog[1,] <- 0

#populate central cell of starting grid on day1
a[3,3,1] <- 1
#set proportion moving
pMove <- 0.6

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove, mnog=mnog)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )
```


### Movement dependent upon vegetation

Within the Hat-trick tsetse simulation flies are assumed to move at different rates in different vegetation types. Flies move more quickly out of cells containing the vegetation types associated with faster movement. In rtsetse this is accomodated by allowing a movement multiplier to be applied to each cell. Where the movement multiplier is greater than 1 flies will move out faster, where it is less than 1 (but still greater than 0) flies will move out more slowly.

We can demonstrate this with a first simple example where a vegetation matrix is created with lower movement in the 2nd row, and the simulation is started with a uniformly populated grid. In the spread maps we can see this row increases in population while those around it decrease.

```{r, "movement-veg1", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#create array to store grids over time
a <- array(0, dim=c(nY, nX, nDays))

#creating empty vegetation matrix
mveg <- array(1, dim=c(nY,nX))
#set 2nd row to lower movement
mveg[2,] <- 0.5

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.6

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove, mveg=mveg)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )
```

In a second example we can increase movement from the 2nd row and in the maps see that the population now becomes lower in that row.

```{r, "movement-veg2", eval=TRUE, echo=FALSE, message=FALSE, fig.width=7}

#create array to store grids over time
a <- array(0, dim=c(nY, nX, nDays))

#creating empty vegetation matrix
mveg <- array(1, dim=c(nY,nX))
#set 2nd row to higher movement
mveg[2,] <- 1.5

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.6

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove, mveg=mveg)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )
```

In a third example we can keep the higher movement in the 2nd row and add lower movement in the 4th row. As expected the population declines in the former and increases in the latter.

```{r, "movement-veg3", eval=TRUE, echo=FALSE, message=FALSE, fig.width=7}
#create array to store grids over time
a <- array(0, dim=c(nY, nX, nDays))

#creating empty vegetation matrix
mveg <- array(1, dim=c(nY,nX))
#set 2nd row to higher movement
mveg[2,] <- 1.5
#set 4th row to lower movement
mveg[4,] <- 0.5

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.6

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove, mveg=mveg)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )
```

Now what happens if we apply this movement to the representation of a real landscape ? Lets first use the Serengeti map dataset stored in rtsetse (copied from Hat-trick).

```{r, "load-and-plot-serengeti", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#get the vegetation map
mVegCats <- rtReadMapVeg( system.file("extdata","vegTanzaniaSerengetiTorr1km.txt", package="rtsetse"))
#plot veg map
rtPlotMapVeg(mVegCats)

```

We can convert the vegetation categories to movement multipliers using the suggestions from Hat-trick for flies preferring Savannah and run the simulation.


```{r, "movement-serengeti1", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#convert vegetation categories to movement multiplier values
#use the same function as mortality, even though it isn't mortality
dfMoveByVeg <-  data.frame(code=c("D","T","O","S","B","G","N"),mortality=c(0.85, 0.9, 0.95, 1, 1.05, 1.1, 0))
#todo change filenames to make this clearer
mveg <- rtSetMortGridFromVeg( sAdultOrPupa ="adult",
                                       mVegetation = mVegCats,
                                       dfMortByVeg = dfMoveByVeg )

#can run for longer
nDays <- 24

#create array to store grids over time
#set dimensions from the veg grid
a <- array(0, dim=c(dim(mveg), nDays))

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.6

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove, mveg=mveg)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )


#to look more closely at the final day
spplot( raster::raster(a[,,nDays]), 
        names.attr=c(paste0("day",nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )

```

The maps above are just for a single age and sex class with movement only and no births and deaths. They give some indication of the population distribution that might be expected if vegetation influenced movement only and there was no effect of vegetation on mortality. Note also that these results do not include any change in behaviour at vegetation boundaries.  


