---
title: "movement in rtsetse"
author: "Andy South"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
#to produce a pdf
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{movement in rtsetse}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

[rtsetse](https://github.com/AndySouth/rtsetse) is an R package that simulates tsetse fly populations on a spatial grid to enable investigation of tsetse control.   

The movement of tsetse flies is an important part of the simulation.   

This vignette demonstrates movement functions of different levels of detail.

First we should load the package after installing from github if necessary.
```{r, eval=TRUE, echo=TRUE, message=FALSE}
#require(devtools)    
#install_github('AndySouth/rtsetse')
require(rtsetse)
```

rtMoveReflectNoGoVeg() simulates movement with reflecting boundaries, avoidance of no-go areas and dependence on vegetation.

This, and the other movement functions in rtsetse, accept a spatial matrix (y,x) defining the population of tsetse flies (of a particular age and gender). The function can be called repeatedly for different ages and sexes, allowing the movement rates to be age and sex dependent. For simplicity the examples shown here will just be for a single age and sex.

### Movement with no vegetation effects or no-go areas

```{r, "movement-no-nogo-or-veg", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#set dimensions of the spatial grid
nY <- 5
nX <- 4
nDays <- 12
#create array to store grids over time
a <- array(0, dim=c(nY, nX, nDays))

#populate central cell of starting grid on day1
a[3,3,1] <- 1
#set proportion moving
pMove <- 0.8

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove)  
}

#quick way of displaying population spread over time
require(raster)
require(sp)
require(RColorBrewer)
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )
```


### Movement avoiding no-go areas

To avoid 'no-go' areas the code is identical to the above with the addition of a matrix for no-go areas. In this example the top row is set as a no-go area. You should be able to see that row remains unoccupied in the spread maps below. 

```{r, "movement-nogo", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#create array to store grids over time
a <- array(0, dim=c(nY, nX, nDays))

#creating empty no-go matrix
mnog <- array(1, dim=c(nY,nX))
#set first row to nogo (0)
mnog[1,] <- 0

#populate central cell of starting grid on day1
a[3,3,1] <- 1
#set proportion moving
pMove <- 0.8

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove, mnog=mnog)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )
```


### Movement dependent upon vegetation

Within the Hat-trick tsetse simulation flies are assumed to move at different rates in different vegetation types. Flies move more quickly out of cells containing the vegetation types associated with faster movement. In rtsetse this is accomodated by allowing a movement multiplier to be applied to each cell. Where the movement multiplier is greater than 1 flies will move out faster, where it is less than 1 (but still greater than 0) flies will move out more slowly.

We can demonstrate this with a first simple example where a vegetation matrix is created with lower movement in the 2nd row, and the simulation is started with a uniformly populated grid. In the spread maps we can see this row increases in population while those around it decrease.

```{r, "movement-veg1", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#create array to store grids over time
a <- array(0, dim=c(nY, nX, nDays))

#creating empty vegetation matrix
mveg <- array(1, dim=c(nY,nX))
#set 2nd row to lower movement
mveg[2,] <- 0.8

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.8

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove, mveg=mveg)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )
```

In a second example we can increase movement from the 2nd row and in the maps see that the population now becomes lower in that row.

```{r, "movement-veg2", eval=TRUE, echo=FALSE, message=FALSE, fig.width=7}

#create array to store grids over time
a <- array(0, dim=c(nY, nX, nDays))

#creating empty vegetation matrix
mveg <- array(1, dim=c(nY,nX))
#set 2nd row to higher movement
mveg[2,] <- 1.2

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.8

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove, mveg=mveg)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )
```

In a third example we can keep the higher movement in the 2nd row and add lower movement in the 4th row. As expected the population declines in the former and increases in the latter.

```{r, "movement-veg3", eval=TRUE, echo=FALSE, message=FALSE, fig.width=7}
#create array to store grids over time
a <- array(0, dim=c(nY, nX, nDays))

#creating empty vegetation matrix
mveg <- array(1, dim=c(nY,nX))
#set 2nd row to higher movement
mveg[2,] <- 1.2
#set 4th row to lower movement
mveg[4,] <- 0.8

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.8

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove, mveg=mveg)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )
```

Now what happens if we apply this movement to the representation of a real landscape ? Lets first use the Serengeti map dataset stored in rtsetse (copied from Hat-trick). The cell size in this map is 1km, 80% of the flies are simulated as moving to one of the four neighbouring cells per day, thus representing a reasonable mean displacement of 800m per day seen in Savannah tsetse (Vale et al. 1984).

```{r, "load-and-plot-serengeti", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#get the vegetation map
mVegCats <- rtReadMapVeg( system.file("extdata","vegTanzaniaSerengetiTorr1km.txt", package="rtsetse"))
#plot veg map
rtPlotMapVeg(mVegCats)

```

We can convert the vegetation categories to movement multipliers using the suggestions from Hat-trick for flies preferring Savannah and run the simulation. Note that the percentage of flies moving out of cells per day seems very high (88% in grass, 80% in savannah and 68% in woodland). Is this reasonable ?   


```{r, "movement-serengeti1-veg", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#convert vegetation categories to movement multiplier values
dfMoveByVeg <-  data.frame(code=c("D","T","O","S","B","G","N"),move=c(0.85, 0.9, 0.95, 1, 1.05, 1.1, 0))
mVeg <- rtSetGridFromVeg( mVegetation=mVegCats, dfLookup=dfMoveByVeg )

#can run for longer
nDays <- 12

#create array to store grids over time
#set dimensions from the veg grid
a <- array(0, dim=c(dim(mVeg), nDays))

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.8

for(day in 2:nDays)
{
  a[,,day] <- rtMoveReflectNoGoVeg(a[,,day-1], pMove=pMove, mveg=mVeg)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )


#to look more closely at the final day
spplot( raster::raster(a[,,nDays]), 
        names.attr=c(paste0("day",nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )

```

The maps above are just for a single age and sex class with movement only and no births and deaths.   


The results above also do not include any change in behaviour at vegetation boundaries. Hat-trick allows for movements to less preferred vegetation types to be decreased. The six potential vegetation types in Hat-trick are ordered by decreasing density of vegetation.

1. Dense Forest
2. Thicket
3. Open Forest
4. Savannah
5. Bush
6. Grass

The default names can be changed, but the ordering by density needs to remain.

The user chooses a preferred ('best') vegetation type for the flies being considered. As expalined in the Hat-trick documentation :

> If the fly is in a certain vegetation type and is headed to the same vegetation type, or to one that is better, ie, the best or a type that is closer to the best type, then the fly will keep going with no adjustment to the evacuation rate. Otherwise, the fly must be headed to a vegetation type that differs more markedly from the best than the one the fly is  in. The fly will then tend to turn back, with a probability that is greater the more the vegetation type to which it is  headed differs from the best. In deciding whether vegetataion is closer to the best, one that is X categories denser than the best is regarded as no worse than one that is X categories less dense than the best.

Movement resulting in a decline in preference of 1,2,3,4,5 categories is set to 30%,10%,3%,1%,0.3% respectively of what it would have been otherwise. The results of implemementing this mechanism in rtsetse are shown below.    

```{r, "movement-serengeti2-boundaries", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#can run for longer
nDays <- 12

#create array to store grids over time
#set dimensions from the veg grid
a <- array(0, dim=c(dim(mVegCats), nDays))

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.8

for(day in 2:nDays)
{
  #because mVeg is not specified here, only the vegetation boundary effect operates
  a[,,day] <- rtMoveReflectNoGoVegBoundary(a[,,day-1], pMove=pMove, mVegCats=mVegCats)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )


#to look more closely at the final day
spplot( raster::raster(a[,,nDays]), 
        names.attr=c(paste0("day",nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )

```

The white cells in the above map show areas of low density that result from areas of less preferred vegetation (grass) neighbouring areas of preferred vegetation (savannah). These boundary cells have lower densities than cells within poor habitat. This is because boundary cells are losing migrants in all directions but are not receiving as many migrants back. This effect persists even after running for thousands of days of movement only, but would be affected by differences in mortality. 


I would have expected that flies should accumulate in the prefferred habitat (e.g. savannah over grassland). That doesn't seem to be happening. Here is a check on the lower right corner of the Serengeti map.

```{r, "movement-serengeti3-boundary-check", eval=TRUE, echo=TRUE, message=FALSE, fig.width=3, fig.show='hold'}
#note reduced fig.width & fig.show='hold' above to fit 2 plots side by side

#get the vegetation map
mVegCats <- rtReadMapVeg( system.file("extdata","vegTanzaniaSerengetiTorr1km.txt", package="rtsetse"))
#subset SE corner
m2 <- mVegCats[40:50,40:50]

nDays <- 99

#create array to store grids over time
#set dimensions from the veg grid
a <- array(0, dim=c(dim(m2), nDays))

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.8

for(day in 2:nDays)
{
  #because mVeg is not specified here, only the vegetation boundary effect operates
  a[,,day] <- rtMoveReflectNoGoVegBoundary(a[,,day-1], pMove=pMove, mVegCats=m2)  
}

#plot vegetation map
rtPlotMapVeg(m2)

#plot final day
spplot( raster::raster(a[,,nDays]), 
        names.attr=c(paste0("day",nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )

```

The figure above shows that even after 99 days the number of flies in grass cells is the same as that in savannah. It is only grass cells bounding savannah that have low numbers. This makes me want to check my mechanism further.

```{r, "movement-serengeti3-boundary-check2", eval=TRUE, echo=TRUE, message=FALSE, fig.width=3, fig.show='hold'}
#note reduced fig.width & fig.show='hold' above to fit 2 plots side by side

#create very simple test veg map
nY <- 5
nX <- 4
# G for Grass, S for Savannah
#vertical split
#m2 <- array("G", dim=c(nY,nX))
#m2[1:(nY*nX/2)] <- "S"
#horizontal split
m2 <- array("G", dim=c(nY,nX))
m2[1:round(nY/2),] <- "S"


nDays <- 99

#create array to store grids over time
#set dimensions from the veg grid
a <- array(0, dim=c(dim(m2), nDays))

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.8

for(day in 2:nDays)
{
  #because mVeg is not specified here, only the vegetation boundary effect operates
  a[,,day] <- rtMoveReflectNoGoVegBoundary(a[,,day-1], pMove=pMove, mVegCats=m2)  
}

#plot vegetation map
rtPlotMapVeg(m2)

#plot final day
spplot( raster::raster(a[,,nDays]), 
        names.attr=c(paste0("day",nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )

```



Adding together the 2 vegetation effects is shown below.

```{r, "movement-serengeti4-both-veg-effects", eval=TRUE, echo=TRUE, message=FALSE, fig.width=7}

#can run for longer
nDays <- 12

#create array to store grids over time
#set dimensions from the veg grid
a <- array(0, dim=c(dim(mVegCats), nDays))

#fill starting grid on day1
a[,,1] <- 1
#set proportion moving
pMove <- 0.8

for(day in 2:nDays)
{
  #because mVeg is not specified here, only the vegetation boundary effect operates
  a[,,day] <- rtMoveReflectNoGoVegBoundary(a[,,day-1], pMove=pMove, mVegCats=mVegCats, mveg=mVeg)  
}

#display population spread over time
spplot( raster::brick(a), 
        names.attr=c(paste0("day",1:nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )


#to look more closely at the final day
spplot( raster::raster(a[,,nDays]), 
        names.attr=c(paste0("day",nDays)), 
        col.regions=c("white",colorRampPalette(brewer.pal(9,"Greens"))(99)) )

```

Now that it seems these vegetation movement effects are working I need to add them into the main simulation including births and deaths.



