---
title: "Report on rtsetse grid model output"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
---

This is a report on the outputs produced from one run of a phase 2 test of the rtsetse model hosted here : https://andysouth.shinyapps.io/shinytse7/ .

See end of the document for the rtsetse code to repeat this run of the model locally.


```{r print args, echo=FALSE}
#removed this because of trickiness in getting the vegetation matrix not to print out
#Parameter values used in this run :

#the arguments and their values passed to rtsetse function
#17/9/14 trying to reimplement this, yes it works
#23/12/14 problem with veg matrix which comes out as long matrix
#print(unlist(lNamedArgsGrid))


#old approach getting list of inputs from Shiny
#but that was a problem once I had multiple navbar tabs
# #getting the named input vars from shiny
# lNamedArgs <- isolate(reactiveValuesToList(input))
# #shinyapps adds extra input vars starting ss-net* or checkbox*
# #I don't want these to appear in the report
# lNamedArgsSub <- lNamedArgs[ substring(names(lNamedArgs),1,3)!='ss-' & substring(names(lNamedArgs),1,8)!='checkbox' ]
# #to order the params alphabetically
# lNamedArgsSub <- lNamedArgsSub[ order(names(lNamedArgsSub)) ]
# #to stop it from using sci notation
# print(unlist(format(lNamedArgsSub,scientific=FALSE)))
```

**Figure 1** Adult population over time summed for the whole grid.
```{r PlotPopGrid, echo=FALSE}
#rtPlotPopGrid(aRecord)

rtPlotPopGrid(v$gridResults)
```


**Figure 2** Map of adults across the grid on final day of the simulation.
```{r PlotMapPop final, echo=FALSE}
#rtPlotMapPop(aRecord, days='final', sex='both')
rtPlotMapPop(v$gridResults, days='final', sex='both')
```


**Figure 3** Maps of adults across the grid over the days of the simulation.
```{r PlotMapPop all, echo=FALSE}
#rtPlotMapPop(aRecord, days='all', ifManyDays='last', sex='both')
rtPlotMapPop(v$gridResults, days='all', ifManyDays='spread', sex='both')
```


**Figure 4** Age structure over time, summed for M&F across the whole grid.
```{r PlotAgeStructure, echo=FALSE}
rtPlotAgeStructure(v$gridResults)
```


**Figure 5** Mean age of adults over time, summed for the whole grid.
```{r PlotMeanAgeGrid, echo=FALSE}
rtPlotMeanAgeGrid(v$gridResults)
```

Code to repeat this run of the model locally :
```{r Code for local run, echo=FALSE}

# string is created in server.r/printParamsGrid
#BEWARE!
#23/12/14 does work if the code tab is selected first, but not yet if not
#i may be able to move the code that creates the string into a function
#that is called from runGridModel()
cat(stringCodeRepeat,"\n")

#cat("the code should be above this ...")

# old way of doing had problems with the veg matrix
# sCommand <- "rt_runGridTestSpread"
# #this creates a vector of 'name=value,'
# vArgs <- paste0(names(lNamedArgsGrid),"=",lNamedArgsGrid,", ")
# #to remove the final comma & space in args list
# vArgs[length(vArgs)] <- substr(vArgs[length(vArgs)],0,nchar(vArgs[length(vArgs)])-2)
# cat( sCommand,"( ",vArgs," )",sep="")

```